The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-04-12 Papers/InProgress/IJCAI/submission.tex
some minor changes for the final version

\title{The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics}
\author{
  \begin{minipage}{0.5\textwidth}\centering
    Christoph Benzm\"uller\thanks{This work was supported by the 
                                  German National Research Foundation (DFG) under
                                  grants BE 2501/9-1 and BE 2501/11-1.} \\
    Freie Universit\"at Berlin, Germany \\
    c.benzmueller@fu-berlin.com
  \end{minipage}
  \begin{minipage}{0.5\textwidth}\centering
    Bruno Woltzenlogel Paleo \\
    Australian National University, Australia \\
    bruno.wp@gmail.com
  \end{minipage}
}

@@

Another novel contribution reported here (in
\S\ref{sec:improvedembedding}) is the implementation of an alternative
embedding for the slightly stronger modal logic \SFiveU, which is
based on a universal accessibility relation. Our experiments have
shown that the new embedding is more efficient, as the following two
previously open problems can now be solved:
\begin{itemize}
\item Automatically proving the final theorem T3 (\textit{Necessarily, there
  exists God}), directly from Scott's \shortcite{ScottNotes} (consistent) axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata).
\item Automatically verifying, in Isabelle/HOL, the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  logical consequences of the argument's axioms.
\end{itemize}

@@

The new embedding of \SFiveU in Isabelle/HOL is shown in
Fig.~\ref{QML_S5U}.
With this improved embedding, the final theorem T3
(\textit{Necessarily, there exists God}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2.5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/LEO-Proof.png}}
\caption{Lines 115--120 of \textsc{Leo-II}'s refutation. Primitive
  substitutions (e.g. with the empty property) are highlighted. In the
red part, property variable $\texttt{SV8}$ has been instantiated with the
$\lambda \texttt{SV16}_{\mu}. \lambda \texttt{SV17}_{\iota}. \bot$, i.e., the (lifted) empty property.
} \label{LEO-Proof}
\end{figure*}

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets).
%
Note that the terms for the empty property\footnote{An additional lambda abstraction occurs in the empty property in LEO-II's proof (and also in the reconstruction in Isabelle) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.} ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with full functional and Boolean extensionality as given
here. Nevertheless, some philosophers\footnote{Private communication with Andr\'e Fuhrmann.} may actually prefer the use of
self-difference over the empty property in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference.

@@

Both the automated detection of the inconsistency in G\"odel's axioms
and the fully automatic proof of T3 from Scott's axioms demonstrate
the potential of our AI technology for philosophy: this technology is,
in its current state of development, already capable of contributing novel results to
metaphysics and to conduct reasoning steps at granularity-levels
beyond common human capabilities.  % ToDo: or even Gödel-level capabilities :-)

% Good improvement (technical sense), Sledgehammer can still be improved
% (\textsc{Leo-II} finds inconsistency when call directly in THF Syntax, but not
% from within Isabelle).

% We also point to several technical challenges which, at least for the
% moment, still restrain non-expert users of these systems from
% independently conducting similar experiments in computational
% metaphysics as reported here and in previous work.

% Technical issues detected within Isabelle-Sledgehammer; room for improvements

%\paragraph*{Acknowledgments} Will be added at a later point. ToDo

@@

\small
\bibliographystyle{named}
\bibliography{Bibliography}

>>>>

\title{The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics}
\author{
  \begin{minipage}{0.5\textwidth}\centering
    Christoph Benzm\"uller\thanks{This work was supported by the 
                                  German National Research Foundation (DFG) under
                                  grants BE 2501/9-1 and BE 2501/11-1.} \\
    Freie Universit\"at Berlin, Germany \\
    \& Stanford University, CSLI, USA \\
    c.benzmueller@gmail.com
  \end{minipage}
  \begin{minipage}{0.5\textwidth}\centering
    Bruno Woltzenlogel Paleo \\
    Australian National University, Australia \\
    bruno.wp@gmail.com
  \end{minipage}
}

@@

Another novel contribution reported here (in
\S\ref{sec:improvedembedding}) is the implementation of an alternative
embedding for the 
% slightly stronger 
more effective modal logic \SFiveU, which is
based on a universal accessibility relation. Our experiments have
shown that the new embedding is more efficient, as the following two
previously open problems can now be solved:
\begin{itemize}
\item Automatically proving the final theorem T3 (\textit{Necessarily, there
  exists God}), directly from Scott's \shortcite{ScottNotes} (consistent) axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata).
\item Automatically verifying, in Isabelle/HOL, the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  logical consequences of the argument's axioms.
\end{itemize}

@@

The new embedding of \SFiveU in Isabelle/HOL is shown in
Fig.~\ref{QML_S5U}.
With this improved embedding, the final theorem T3
(\textit{Necessarily, there exists God}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2.5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{leo2} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

\begin{figure*}
% \begin{minipage}{.01\textwidth}\small
% \vskip2cm * 
% \end{minipage}
% \begin{minipage}{.99\textwidth}
% {\includegraphics[width=\textwidth]{./Images/LEO-Proof.png}}
% \end{minipage}
\centerline{\includegraphics[width=\textwidth]{./Images/LEO-Proof.png}}
\caption{Lines 115--120 of \textsc{Leo-II}'s refutation. Primitive
  substitutions (e.g. with the empty property) are highlighted. In the
red part (see $\boldsymbol{\leftarrow}$), property variable $\texttt{SV8}$ has been instantiated with the
$\lambda \texttt{SV16}_{\mu}. \lambda \texttt{SV17}_{\iota}. \bot$, i.e., the (lifted) empty property.
} \label{LEO-Proof}
\end{figure*}

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets).
%
Note that the terms for the empty property\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and also in the reconstruction in Isabelle) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.} ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with full functional and Boolean extensionality as given
here. Nevertheless, some philosophers\footnote{Private communication with Andr\'e Fuhrmann.} may actually prefer the use of
self-difference over the empty property in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference.

@@

Both the automated detection of the inconsistency in G\"odel's axioms
and the fully automatic proof of T3 from Scott's axioms demonstrate
the potential of our AI technology for philosophy: this technology is,
in its current state of development, already capable of contributing novel results to
metaphysics and to conduct reasoning steps at granularity-levels
beyond common human capabilities.  
\vfill

\noindent\textbf{Acknowledgments:} We thank Chad Brown, who contributed to the rational  reconstruction
of the inconsistency argument.

\pagebreak
%A discussion between Chad Brown and Benzm\"uller
% significantly influenced the rational reconstruction of the inconsistency reported by \textsc{LEO-II}.

@@

\small
\bibliographystyle{named}
%\bibliography{Bibliography}

\begin{thebibliography}{}

\bibitem[\protect\citeauthoryear{Adams}{1995}]{Adams}
R.M. Adams.
\newblock Introductory note to *1970.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpubl. Essays and
  Letters}}. Oxford Univ. Press, 1995.

\bibitem[\protect\citeauthoryear{Anderson and
  Gettings}{1996}]{AndersonGettings}
A.C. Anderson and M.~Gettings.
\newblock {G\"odel} ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}, pages 167--172. {Springer},
  1996.

\bibitem[\protect\citeauthoryear{Anderson}{1990}]{Anderson}
C.A. Anderson.
\newblock Some emendations of {G{\"o}del's} ontological proof.
\newblock {\em Faith and Philosophy}, 7(3), 1990.

\bibitem[\protect\citeauthoryear{Andrews}{2014}]{andrewsSEP}
P.B. Andrews.
\newblock Church's type theory.
\newblock In E.N. Zalta, editor, {\em The Stanford Encyclopedia of Philosophy}.
  Spring 2014 edition, 2014.

\bibitem[\protect\citeauthoryear{Anselm}{1078}]{Proslogion}
St. Anselm.
\newblock Proslogion.
\newblock In M.~Charlesworth, editor, {\em St.~Anselm's Proslogion}.
  Oxford:OUP, 1078.
\newblock Republished in 1965.

\bibitem[\protect\citeauthoryear{Benzm{\"u}ller and Paulson}{2013}]{J23}
C.~Benzm{\"u}ller and L.C. Paulson.
\newblock Quantified multimodal logics in simple type theory.
\newblock {\em Logica Universalis}, 7(1):7--20, 2013.

\bibitem[\protect\citeauthoryear{Benzm{\"u}ller and
  Woltzenlogel-Paleo}{2013a}]{J30}
C.~Benzm{\"u}ller and B.~Woltzenlogel-Paleo.
\newblock Formalization, mechanization and automation of {G{\"o}del's} proof of
  {God's} existence.
\newblock {\em arXiv:1308.4526}, 2013.
\newblock Preprint available as arXiv:1308.4526.

\bibitem[\protect\citeauthoryear{Benzm\"uller and
  Woltzenlogel-Paleo}{2013b}]{J28}
C.~Benzm\"uller and B.~Woltzenlogel-Paleo.
\newblock {G{\"o}del's God in Isabelle/HOL}.
\newblock {\em Archive of Formal Proofs}, 2013, 2013.

\bibitem[\protect\citeauthoryear{Benzm{\"u}ller and
  Woltzenlogel-Paleo}{2014}]{C40}
C.~Benzm{\"u}ller and B.~Woltzenlogel-Paleo.
\newblock Automating {G\"{o}del's} ontological proof of {God}'s existence with
  higher-order automated theorem provers.
\newblock In 
  {\em ECAI 2014}, volume 263 of {\em Frontiers in Artificial Intelligence and
  Applications}, pages 93 -- 98. IOS Press, 2014.

\bibitem[\protect\citeauthoryear{Benzm{\"u}ller \bgroup \em et al.\egroup
  }{2015}]{leo2}
Christoph Benzm{\"u}ller, Lawrence~C. Paulson, Nik Sultana, and Frank
  Thei{\ss}.
\newblock The higher-order prover {LEO-II}.
\newblock {\em J. of Automated Reasoning}, 55(4):389--404, 2015.

\bibitem[\protect\citeauthoryear{Bj{\o}rdal}{1999}]{Bjordal}
F.~Bj{\o}rdal.
\newblock Understanding {G\"{o}del’s} ontological argument.
\newblock In T.~Childers, editor, {\em The Logica Yearbook 1998}. Filosofia,
  1999.

\bibitem[\protect\citeauthoryear{Blackburn \bgroup \em et al.\egroup
  }{2001}]{Blackburn}
P.~Blackburn, M.~de~Rijke, and Y.~Venema.
\newblock {\em Modal Logic}.
\newblock Cambridge University Press, 2001.

\bibitem[\protect\citeauthoryear{Blanchette and Nipkow}{2010}]{Nitpick}
J.C. Blanchette and T.~Nipkow.
\newblock Nitpick: A counterexample generator for higher-order logic based on a
  relational model finder.
\newblock In {\em ITP 2010}, number 6172 in LNCS, pages 131--146. Springer,
  2010.

\bibitem[\protect\citeauthoryear{Blanchette \bgroup \em et al.\egroup
  }{2013}]{Sledgehammer}
J.C. Blanchette, S.~B\"ohme, and L.C. Paulson.
\newblock Extending {Sledgehammer} with {SMT} solvers.
\newblock {\em J. of Automated Reasoning}, 51(1):109--128, 2013.

\bibitem[\protect\citeauthoryear{Brown}{2012}]{Satallax}
C.E. Brown.
\newblock Satallax: An automated higher-order prover.
\newblock In {\em IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117. Springer,
  2012.

\bibitem[\protect\citeauthoryear{Fitelson and Zalta}{2007}]{FitelsonZalta}
Branden Fitelson and Edward~N. Zalta.
\newblock Steps toward a computational metaphysics.
\newblock {\em J. Philosophical Logic}, 36(2):227--247, 2007.

\bibitem[\protect\citeauthoryear{Fuhrmann}{2016}]{fuhrmann15:_blogg_goedel}
A.~Fuhrmann.
\newblock Blogging {G\"odel}: His ontological argument in the public eye.
\newblock In K.~\'{S}wi{e}torzecka, editor, {\em Forthcoming in
  \textit{G\"odel’s Ontological Argument - History, Modifications, and
  Controversies}}. 2016.

\bibitem[\protect\citeauthoryear{G\"odel}{1970}]{GoedelNotes}
K.~G\"odel.
\newblock {\em Appx. A: Notes in Kurt G\"odel's Hand}, pages 144--145.
\newblock In Sobel \shortcite{sobel2004logic}, 1970.

\bibitem[\protect\citeauthoryear{H\'ajek}{1996}]{Hajek1}
P.~H\'ajek.
\newblock Magari and others on {G\"odel’s} ontological proof.
\newblock In A.~Ursini and P.~Agliano, editors, {\em Logic and algebra}, page
  125–135. Dekker, New York etc., 1996.

\bibitem[\protect\citeauthoryear{H\'ajek}{2001}]{Hajek2}
P.~H\'ajek.
\newblock {Der Mathematiker und die Frage der Existenz Gottes}.
\newblock In B.~Buldt~et al., editor, {\em {Kurt G{\"o}del. Wahrheit und
  Beweisbarkeit}}, pages 325--336. öbv \& hpt, Wien, 2001.
\newblock ISBN 3-209-03835-X.

\bibitem[\protect\citeauthoryear{H{\'{a}}jek}{2002}]{Hajek3}
P.~H{\'{a}}jek.
\newblock A new small emendation of {G{\"{o}}del's} ontological proof.
\newblock {\em Studia Logica}, 71(2):149--164, 2002.

\bibitem[\protect\citeauthoryear{Hazen}{1998}]{Hazen}
A.P. Hazen.
\newblock On {G\"odel's} ontological proof.
\newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

\bibitem[\protect\citeauthoryear{McCune}{2010}]{prover9-mace4}
W.~McCune.
\newblock {Prover9} and {Mace4} (2005--2010).
\newblock {\small \verb|http://www.cs.unm.edu/~mccune/prover9/|}, 2010.

\bibitem[\protect\citeauthoryear{Muskens}{2006}]{homl}
R.~Muskens.
\newblock {Higher Order Modal Logic}.
\newblock In P.~Blackburn~et al., editor, {\em Handbook of Modal Logic},
  Studies in Logic and Practical Reasoning, pages 621--653. Elsevier,
  Dordrecht, 2006.

\bibitem[\protect\citeauthoryear{Nipkow \bgroup \em et al.\egroup
  }{2002}]{NPW02}
T.~Nipkow, L.~Paulson, and M.~Wenzel.
\newblock {\em {Isabelle/HOL: A Proof Assistant for Higher-Order Logic}}.
\newblock Number 2283 in LNCS. Springer, 2002.

\bibitem[\protect\citeauthoryear{Ohlbach}{1991}]{DBLP:journals/logcom/Ohlbach91}
H.J. Ohlbach.
\newblock Semantics-based translation methods for modal logics.
\newblock {\em J. Log. Comput.}, 1(5):691--746, 1991.

\bibitem[\protect\citeauthoryear{Oppenheimer and Zalta}{2011}]{oppenheimer11}
P.E. Oppenheimer and E.N. Zalta.
\newblock A computationally-discovered simplification of the ontological
  argument.
\newblock {\em Australasian J. of Philosophy}, 89(2):333--349, 2011.

\bibitem[\protect\citeauthoryear{Oppy}{1996}]{oppy96:_goedel_ontol_argum}
G.~Oppy.
\newblock G{\"o}delian ontological arguments.
\newblock {\em Analysis}, 56(4):226--230, 1996.

\bibitem[\protect\citeauthoryear{Oppy}{2000}]{oppy00:_respon_gettin}
G.~Oppy.
\newblock Response to {Gettings}.
\newblock {\em Analysis}, 60(4):363--367, 2000.

\bibitem[\protect\citeauthoryear{Oppy}{2008}]{oppy08:_higher_order_ontol_argum}
G.~Oppy.
\newblock Higher-order ontological arguments.
\newblock {\em Philosophy Compass}, 3(5):1066--1078, 2008.

\bibitem[\protect\citeauthoryear{Oppy}{2015}]{sep-ontological-arguments}
G.~Oppy.
\newblock Ontological arguments.
\newblock In Edward~N. Zalta, editor, {\em The Stanford Encyclopedia of
  Philosophy}. Spring 2015 edition, 2015.

\bibitem[\protect\citeauthoryear{Owre \bgroup \em et al.\egroup
  }{1992}]{cade92-pvs}
{S.} Owre, {J.}~{M.} Rushby, and {N.} Shankar.
\newblock {PVS:} {A} prototype verification system.
\newblock In Deepak Kapur, editor, {\em CADE}, volume 607 of {\em LNAI}, pages
  748--752, Saratoga, {NY}, jun 1992. Springer.

\bibitem[\protect\citeauthoryear{Rushby}{2013}]{rushby13}
J.~Rushby.
\newblock The ontological argument in {PVS}.
\newblock In {\em Proc.~of CAV Workshop ``Fun With Formal Methods''}, St.
  Petersburg, Russia, 2013.

\bibitem[\protect\citeauthoryear{Scott}{1972}]{ScottNotes}
D.~Scott.
\newblock {\em Appx. B: Notes in Dana Scott's Hand}, pages 145--146.
\newblock In Sobel \shortcite{sobel2004logic}, 1972.

\bibitem[\protect\citeauthoryear{Sobel}{1987}]{Sobel}
J.H. Sobel.
\newblock G\"odel's ontological proof.
\newblock In {\em {On Being and Saying. Essays for Richard Cartwright}}, pages
  241--261. {MIT Press}, 1987.

\bibitem[\protect\citeauthoryear{Sobel}{2004}]{sobel2004logic}
J.H. Sobel.
\newblock {\em Logic and Theism: Arguments for and Against Beliefs in God}.
\newblock Cambridge U. Press, 2004.

\bibitem[\protect\citeauthoryear{Sultana and Benzm{\"u}ller}{2013}]{W47}
N.~Sultana and C.~Benzm{\"u}ller.
\newblock Understanding {LEO-II's} proofs.
\newblock In Konstantin Korovin, Stephan Schulz, and Eugenia Ternovska,
  editors, {\em IWIL 2012}, volume~22 of {\em EPiC Series}, pages 33--52,
  Merida, Venezuela, 2013. EasyChair.

\bibitem[\protect\citeauthoryear{Sutcliffe and Benzm{\"u}ller}{2010}]{J22}
G.~Sutcliffe and C.~Benzm{\"u}ller.
\newblock Automated reasoning in higher-order logic using the {TPTP THF}
  infrastructure.
\newblock {\em J. of Formalized Reasoning}, 3(1):1--27, 2010.

\bibitem[\protect\citeauthoryear{Williamson}{2013}]{williamson13}
T.~Williamson.
\newblock {\em Modal Logic as Metaphysics}.
\newblock Oxford:OUP, 2013.

\end{thebibliography}
